Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-09 paper.tex
terminado capitulo de usos da ops

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
\newcommand{\opa}{\textsc{op}a}
\newcommand{\ocd}{\textsc{ocd}}
\newcommand{\ontologiaa}{\textsc{o}ntologiaa}
\newcommand{\obs}{\textsc{obs}}
\newcommand{\pubby}{\textsc{p}ubby}
\newcommand{\rdf}{\textsc{rdf}}
\newcommand{\mysql}{\textsc{m}y\textsc{sql}}
\newcommand{\aan}{\textsc{aa}}
\newcommand{\cidadedemocratica}{\textsc{c}idade \textsc{d}emocr\'atica}
\newcommand{\participa}{\textsc{p}articipa.br}
\newcommand{\ontop}{\textsc{o}n\textsc{T}op}
\newcommand{\quest}{\textsc{Q}uest}
\newcommand{\obda}{\textsc{obda}}

@@

Some examples of this usage can be given by SparQL queries and concise explanations:
\begin{itemize}
    \item \texttt{"select ?s ?s2 ?s3 where \{?s a ops:SocialActor . ?s2 a ops:Person . ?s3 a ops:Organization\}"}: this query retrieves all social actors (returned in variable \texttt{?s}), be each a person or an organization; retrieves all persons (variable \texttt{?s2}); and retrieves all organizations (\texttt{?s3}). In a similar manner, one can retrieve all roles played, all executers, all initiators and all supporters.
    \item \texttt{"select ?s ?o where \{?s ops:starts ?o\}"}: this query retrieves all causes ({\tt ?o}) and their initiators ({\tt ?s}).
    \item \texttt{"select ?s ?s2 ?o ?o2 where \{?s a ops:Action . ?s ops:belongsTo ?o . ?s2 ops:executes ?s . ?s ops:produces ?o2\}"}: this query retrieves all actions (\texttt{?s}) along their Action Field (\texttt{?o}), their Executer (\texttt{?s2}) and their Results (\texttt{?o2}).
\end{itemize}

Noteworthy is that while {\tt opa:Participant} can be used to retrieve all \participa\ participants, {\tt ocd:Participant} can be used to retrieve all \cidadedemocratica\ participants, and {\tt aa:User} can be used to retrieve all \aan\ participants, their upper ontology class {\tt opa:Participant} retrieves all of them and relates these entities directly to the class of generic actors of social participation processes.

@@

%\begin{landscape}
%\begin{figure}
\begin{sidewaysfigure*}
    \vspace{24.5em}
    \hspace*{-3.5em}
        \includegraphics[width=1.05\textwidth]{figs/opsExpandedTax}
    \caption{A taxonomic diagram of an expanded instance of OSP. Darker orange marks defined classes: PaidExecutor is defined as being the subject of a \texttt{receivesFrom} relation with a \texttt{SocialActor}; \texttt{DownloadedMob} is defined by being the subject of a \texttt{convoquedBy} relation with a \texttt{SocialNetwork}. OWL code is online for live editing~\cite{owlExp}. The new classes added to OSP are in Table~\ref{ospFooClass}}
    \label{fig:owlExp}
\end{sidewaysfigure*}
%\end{figure}
%\end{landscape}
To this point, OSP is complete (no missing classes, properties and restrictions with respect to documentation), correct (no spaces in URIs and \texttt{min 1} restrictions specified as existential restrictions), and names of classes and properties standardized (English in CamelCase and headlessCamelCase). This way, OSP matches, with coherence, CVSP community online documentation~\cite{corais}. As no greater consensus was reached, further developments of OPA should wait feedback and unroll of community processes.

As an example of possible aditional classes, another instance of the ontology is uploaded to Webprotege, from which some newly created classes and properties are exhibited~\cite{owlExp}. Table~\ref{ospFooClass} is dedicated to these aditional classes while Figure~\ref{fig:owlExp} exhibits the resulting taxonomic structure.

The property \texttt{receivesFrom} was also added and has an inverse: \texttt{SocialActor ``paysTo'' Executor}. Also, the \texttt{DownloadedMod} class is a defined class by the restricton: \texttt{Mob convoquedBy Network}, with a newly defined property \texttt{convoquedBy}.

This is one of the numerous ways by which OSP might deal with further classes, properties and restrictions. This particular expansion was chosen as an example by plain observance of community documentation and recent social affairs, such as the Brazilian protests.

\begin{table*}
  \centering
  \caption{New classes considered for expansion of the OSP. The taxonomic organization of these classes, within OSP can be oberved in Figure~\ref{fig:owlExp}. Further information is in Section~\ref{downwards}.}
  \begin{tabular}{|l|l|p{5cm}|p{5cm}|}\hline
{\bf new class} & {\bf subclass of} & {\bf description} & {\bf further notes} \\\hline\hline
SocialNetwork & Organization & a social structure made up of social actors (such as individuals or organizations) and a set of dyadic ties between these actors & --//--\\\hline
FreeScaleNetwork & SocialNetwork & a network whoose connectivity follows a power law & disjoint with UniformrandomNetwork and GeographicNetwork \\ \hline
UniformRandomNetwork & SocialNetwork & also known as Erd\"os Renyi network, this network sets, with equal propability, an edge between each pair of nodes & disjoint with FreeScaleNetwork and GeographicNetwork\\\hline
GeographicNetwork & SocialNetwork & a network whose connectivity is related to the distance of nodes in a metric space & Disjoint with both FreeScaleNetwork and UniformRandomNetwork \\\hline
SmallWorldNetwork & SocialNetwork & a network where most nodes can be reached from other nodes with few hops or steps & not disjoint with FreeScaleNetwork, UniformNetwork and GeographicNetwork\\\hline\hline
InformalOrganization & Organization & an organization that is not formalized & disjoint with Intitution \\ \hline
Mob & InformalOrganization & a crowd of individuals & --//-- \\\hline
GiantMob & Mob & a crowd with more than 10,000 individuals & --//--\\\hline
DownloadedMob & Mob & a Mob convoqued by a network & this is a defined class, by being a network and being the subject of the relation convoquedBy with object Network \\ \hline\hline
Institution & Organization & a mechanism of social order that governs a set of individuals & disjoint with InformalOrganization \\\hline
PublicInstitution & Institution & an institution backed through public funds and controlled by the state & disjoint with Private Institution \\ \hline
PrivateInstitution & Institution & an institution backed through private fundings and controlled by private parties & disjoint with PublicInstitution \\ \hline
AcademicInstitution & Institution & an institution dedicated to education and research, which grants academic degrees & --//-- \\ \hline
NGO & Institution & a legally consituted corporation created by natural or legal people that operate independently from anu form of government & --//-- \\\hline
SpuriousInstitution & Institution & an institution that holds prominent illegitimate or corrupt characteristcs & --//-- \\\hline
ExoticInstitution & Institution & and institution that does not fir previous classes or is characterized by a very unique traces & --//-- \\\hline\hline
VoluntaryExecutor & Executor & an executor that receives no formal reward for the tasks & disjoint with PaidExecutor \\ \hline
PaidExecutor & Executor & an Executor that receives formal reward for the tasks accomplished & a defined class, being an Executor and the subjecto of a receivesFrom predicate with SocialActor as object\\ \hline
  \end{tabular}
  \label{ospFooClass}
\end{table*}

@@

OSP usage might not be obvious at first. How is data being linked? How is field knowledge being organized? Core principles of this utility, reflecting the exposition in Section~\ref{sec:into}, can be understood by the following observations:
\begin{itemize}
    \item Different participation instances have social actors, actions being developed, organizations involved, problems being tackled, etc. These can yield one consistent database by means of OSP usage.
    \item One can understand the mutually exclusive nature of being a paid or a voluntary contributor by observing the expanded version of OSP, exposed above. Also, noticing the fact that a mob can be very big or not, and that it can be convoqued or not by a Network, can make the field more neat for a newcommer or ease discussions and problematizations for senior researchers or politicians.
\end{itemize}

Imagined examples are useful to make these points clear:
\begin{itemize}
    \item Suppose a public SparQL endpoint unifies several participation instances by means of OSP. Thus, the number of participants (Social Actors) is publicly available. Also, depending on the platforms, one can observe how many are individuals, how many are organizations, and understand to which extent the corporative influence is explicit. One can observe how many of the participants are the same in each platform, and what roles they take, and make conclusions about how much society is really participating or if these processes are manipulated by the same agents. One can also gaze uppon the problems being discussed and which solutions are being proposed, therefore easing the sense of what is being considered important and valid as public discussions. This list of possibilities is endless, specially when OSP variations and expansions are considered.
    \item Suppose a person, say Jessica, has a new proposal for a participative system that uses OSP. She can have a quite concise understanding of the concepts involved, and how they relate, by means of OSP. She can now make very objective observations and deliver clear suggestions that relate directly to the system being used. She can even make an OSP variation or another ontology, as a way to confront the schemes.
    \item Supose there is an automatic system for exhibiting indicators about social participation (how effective it has been, how wide the scope of interests, etc). Instances that are integrated by OSP are queried for information and, for example, this system registers any organization involved as a social actor. In the case of the expanded OSP exposed above, for example, the system registers any mob, whose incidence was recorded in the database as a \texttt{DownloadedMob}, as related to some social network, be the network known or not.
\end{itemize}

\subsection{Real use cases}\label{examplesUsage}

@@

\newpage
%\nocite{*}
\bibliography{paper}% Produces the bibliography via BibTeX.

>>>>

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
\newcommand{\opa}{\textsc{op}a}
\newcommand{\ocd}{\textsc{ocd}}
\newcommand{\ontologiaa}{\textsc{o}ntologiaa}
\newcommand{\obs}{\textsc{obs}}
\newcommand{\pubby}{\textsc{p}ubby}
\newcommand{\rdf}{\textsc{rdf}}
\newcommand{\mysql}{\textsc{m}y\textsc{sql}}
\newcommand{\aan}{\textsc{aa}}
\newcommand{\cidadedemocratica}{\textsc{c}idade \textsc{d}emocr\'atica}
\newcommand{\participa}{\textsc{p}articipa.br}
\newcommand{\ontop}{\textsc{o}n\textsc{T}op}
\newcommand{\quest}{\textsc{Q}uest}
\newcommand{\obda}{\textsc{obda}}
\newcommand{\pnud}{\textsc{pnud}}
\newcommand{\onu}{\textsc{onu}}
\newcommand{\serpro}{\textsc{S}erpro}
% triplification of data

@@

Some examples of this usage can be given by \sparql\ queries and concise explanations:
\begin{itemize}
    \item \texttt{"select ?s ?s2 ?s3 where \{?s a ops:SocialActor . ?s2 a ops:Person . ?s3 a ops:Organization\}"}: this query retrieves all social actors (returned in variable \texttt{?s}), be each a person or an organization; retrieves all persons (variable \texttt{?s2}); and retrieves all organizations (\texttt{?s3}). In a similar manner, one can retrieve all roles played, all executers, all initiators and all supporters.
    \item \texttt{"select ?s ?o where \{?s ops:starts ?o\}"}: this query retrieves all causes ({\tt ?o}) and their initiators ({\tt ?s}).
    \item \texttt{"select ?s ?s2 ?o ?o2 where \{?s a ops:Action . ?s ops:belongsTo ?o . ?s2 ops:executes ?s . ?s ops:produces ?o2\}"}: this query retrieves all actions (\texttt{?s}) along their Action Field (\texttt{?o}), their Executer (\texttt{?s2}) and their Results (\texttt{?o2}).
\end{itemize}

Noteworthy is that while {\tt opa:Participant} can be used to retrieve all \participa\ participants, {\tt ocd:Participant} can be used to retrieve all \cidadedemocratica\ participants, and {\tt aa:User} can be used to retrieve all \aan\ participants; their upper ontology class {\tt ops:SocialActor} retrieves all of them and relates these entities directly to the class of generic actors of social participation processes.

@@

\begin{sidewaysfigure*}
    \vspace{24.5em}
    \hspace*{-3.5em}
        \includegraphics[width=1.05\textwidth]{figs/opsExpandedTax}
    \caption{A taxonomic diagram of an expanded instance of OSP. Darker orange marks defined classes: PaidExecutor is defined as being the subject of a \texttt{receivesFrom} relation with a \texttt{SocialActor}; \texttt{DownloadedMob} is defined by being the subject of a \texttt{convoquedBy} relation with a \texttt{SocialNetwork}. OWL code is online for live editing~\cite{owlExp}. The new classes added to OSP are in Table~\ref{ospFooClass}}
    \label{fig:owlExp}
\end{sidewaysfigure*}

\opsi\ matches, \vcps\ online documentation~\cite{corais}. An example of additional classes, an expanded \ops\ ontology is exposed in this section and is uploaded to Webprotege~\cite{owlExp}. Table~\ref{ospFooClass} is dedicated to these additional classes while Figure~\ref{fig:owlExp} presents the resulting taxonomic structure.

The property {\tt ops:receivesFrom} was also added and has an inverse: \texttt{ops:SocialActor ops:paysTo ops:Executor}. Also, the {\tt ops:DownloadedMod} class is a defined class by the restriction: \texttt{ops:Mob ops:convoquedBy ops:Network}, with a newly defined property \texttt{ops:convoquedBy}.

This is one of the numerous ways by which \ops\ might deal with further classes, properties and restrictions. This particular expansion was chosen as an example by direct observance of \vcps\ documentation and recent social affairs, such as the Brazilian protests.

\begin{table*}
  \centering
  \caption{New classes considered for expansion of the \ops. The taxonomic organization of these classes, within \ops\ can be observed in Figure~\ref{fig:owlExp}. Further information is in Section~\ref{downwards}.}
  \begin{tabular}{|l|l|p{5cm}|p{5cm}|}\hline
{\bf new class} & {\bf subclass of} & {\bf description} & {\bf further notes} \\\hline\hline
SocialNetwork & Organization & a social structure made up of social actors (such as individuals or organizations) and a set of dyadic ties between these actors & --//--\\\hline
FreeScaleNetwork & SocialNetwork & a network whoose connectivity follows a power law & disjoint with UniformrandomNetwork and GeographicNetwork \\ \hline
UniformRandomNetwork & SocialNetwork & also known as Erd\"os Renyi network, this network sets, with equal propability, an edge between each pair of nodes & disjoint with FreeScaleNetwork and GeographicNetwork\\\hline
GeographicNetwork & SocialNetwork & a network whose connectivity is related to the distance of nodes in a metric space & Disjoint with both FreeScaleNetwork and UniformRandomNetwork \\\hline
SmallWorldNetwork & SocialNetwork & a network where most nodes can be reached from other nodes with few hops or steps & not disjoint with FreeScaleNetwork, UniformNetwork and GeographicNetwork\\\hline\hline
InformalOrganization & Organization & an organization that is not formalized & disjoint with Intitution \\ \hline
Mob & InformalOrganization & a crowd of individuals & --//-- \\\hline
GiantMob & Mob & a crowd with more than 10,000 individuals & --//--\\\hline
DownloadedMob & Mob & a Mob convoqued by a network & this is a defined class, by being a network and being the subject of the relation convoquedBy with object Network \\ \hline\hline
Institution & Organization & a mechanism of social order that governs a set of individuals & disjoint with InformalOrganization \\\hline
PublicInstitution & Institution & an institution backed through public funds and controlled by the state & disjoint with Private Institution \\ \hline
PrivateInstitution & Institution & an institution backed through private fundings and controlled by private parties & disjoint with PublicInstitution \\ \hline
AcademicInstitution & Institution & an institution dedicated to education and research, which grants academic degrees & --//-- \\ \hline
NGO & Institution & a legally consituted corporation created by natural or legal people that operate independently from anu form of government & --//-- \\\hline
SpuriousInstitution & Institution & an institution that holds prominent illegitimate or corrupt characteristcs & --//-- \\\hline
ExoticInstitution & Institution & and institution that does not fir previous classes or is characterized by a very unique traces & --//-- \\\hline\hline
VoluntaryExecutor & Executor & an executor that receives no formal reward for the tasks & disjoint with PaidExecutor \\ \hline
PaidExecutor & Executor & an Executor that receives formal reward for the tasks accomplished & a defined class, being an Executor and the subjecto of a receivesFrom predicate with SocialActor as object\\ \hline
  \end{tabular}
  \label{ospFooClass}
\end{table*}

@@

\opsi\ usage might not be obvious at first. How is data being linked? How is field knowledge being organized? Why and by whom? Core principles of this \ops\ utilities, can be understood by the following observations:
\begin{itemize}
    \item Different participation instances have social actors, actions being developed, organizations involved, problems being tackled, etc. These can yield one consistent database by means of \ops\ usage.
    \item One can understand and share the mutually exclusive nature of being a paid or a voluntary contributor by observing the expanded version of \ops\ (see Section~\ref{downwards}. Also, noticing the fact that a mob can be very big or not, and that it can be convoked or not by a Network, can make the field more neat for a newcomer or ease discussions and problematization for senior researchers or politicians.
    \item Other fields of human knowledge and practice also have agents, problems and so forth. These can be linked to participatory data and ontologies by means of \ops.
\end{itemize}

Fictional examples are useful to make these points clear:
\begin{itemize}
    \item Suppose a public \sparql\ endpoint unifies several participation instances by means of \ops\ (we will see in Section~\ref{sec:real} that this is not really fictional). Thus, the number of participants is publicly available ({\tt |ops:SocialActor|}). Also, depending on the platforms involved, one can observe how many of these participants are individuals ({\tt ops:Person}), how many are organizations ({\tt ops:Organization}), and understand to which extent the corporative influence is explicit. One can observe how many of the participants are the same in each platform, and what roles they take, and make conclusions about how much the society is really participating or if these processes are manipulated by a few agents ({\tt ops:SocialActor}). One can also gaze uppon the problems being discussed and which solutions are being proposed, therefore easing the sense of what is being considered important and valid as public discussions. This list of possibilities is endless, specially when \ops\ variations and expansions are considered.
    \item Suppose a person, say Jessica, has a new proposal for a participative system that uses \ops. She can have a quite concise understanding of the concepts involved, and how they relate. She can now make very objective observations and deliver clear suggestions that relate directly to the systems used or envisioned. She can make an \ops\ variation or another ontology, as a way to confront paradigms.
    \item Suppose there is a system for exhibiting indicators about social participation (how effective it has been, how wide the scope of interests, etc.). Instances that are integrated by \ops\ can be queried for information and, for example, this system registers any organization involved as a social actor. Also, reflecting the expanded \ops\ exposed above, the system registers any mob involved, whose incidence was recorded in the database as a \texttt{DownloadedMob}, as related to some social network (the network might be unknown).
\end{itemize}

\subsection{Real use cases}\label{sec:real}
\opsi\ is a recent ontology. Even so, some real use cases can be pointed, from which are most notable:
\begin{itemize}
        \item The \pnud/\onu\ consultant contract 2013/00056, project BRA/12/018, was heavily influenced by \ops. Within written products delivered are other participatory ontologies, such as \opa, \ocd, \ontologiaa\ and \obs, which relates directly to \ops. Also, some methods for analysing \ops\ related data and for resource recommendation were delivered. These developments where done by the computational the physics researcher Renato Fabbri (main author of the present article), in collaboration with other parties, specially the Brazilian General Secretariat of the Republic, University of Bras\'ilia researchers, and free software parties. Data from three participatory instances were triplified: \participa, \aa and \cidadedemocratica, these are also related to \ops.
    \item Another \pnud/\onu\ contract was responsible for some advances in information technologies. A special case is a dedicated \ops\ expansion\cite{opsFernando}.
    \item The Brazilian Federal Data Processing Service (\serpro) is triplifying \participa\ data (the Brazilian federal social participation portal). Is this process, they as making direct use of \ops~\cite{tripSerpro}.
    \item The linked data Brazilian community had some maturing related to \ops\ beyond the ontologies developed, \pnud/\onu documentation and \serpro\ triplification, and \vcps. Interested government, civil society and academic parties circulated \ops\ documentation for conceptual and technological goals~\cite{eventoParticipativo,documentacaoParticipativa}.
\end{itemize}

@@

\section{Glossary of terms, achronims and abbreviations}\label{ap:glossary}
\newpage
%\nocite{*}
\bibliography{paper}% Produces the bibliography via BibTeX.
